---
title: "Compare PCA"
author: "Davide Risso"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="markup")
```

```{r packages, message=FALSE, warning=FALSE}
library(TENxBrainData)
library(SingleCellExperiment)
library(BiocSingular)
library(BiocParallel)
library(scuttle)

ncores <- 10
```

# Sparse HDF5 file (TENxMatrix)

First, we read the data in sparse format from the 10X Genomics file.

```{r readin}
tenx <- TENxMatrix("1M_neurons_filtered_gene_bc_matrices_h5.h5")
class(tenx)

sce <- SingleCellExperiment(assays = list(counts=tenx))
```

Normalization is breaking the sparsity (why?) so for now we only use log transformation.

```{r log}
# this destroys sparsity... doing something simpler for now
# system.time(sce <- computeLibraryFactors(sce, BPPARAM = BiocParallel::MulticoreParam(6)))
# system.time(sce <- logNormCounts(sce, BPPARAM = BiocParallel::MulticoreParam(6)))
assay(sce, "logcounts") <- log1p(counts(sce))
```

We select the top variable genes. Because of R limits, rowVars does not work on the full matrix, we use a random subset of cell to compute the variances.

```{r var}
## rowVars does not work with so many cells, using a random subset for now
set.seed(134)
idx <- sample(seq_len(ncol(sce)), 1e4)
idx <- sort(idx)
system.time(vars <- rowVars(logcounts(sce[,idx])))
names(vars) <- rownames(sce)
vars <- sort(vars, decreasing = TRUE)

geneidx <- which(rownames(sce) %in% names(vars)[1:1000])
top <- sce[geneidx,]
is_sparse(logcounts(top))
```

Log and subset are delayed operations, we need to realize the matrix to be able to run PCA.

```{r realize}
ltop <- realize(logcounts(top), BACKEND = "TENxMatrix")
ltop <- realize(t(ltop), BACKEND = "TENxMatrix")
seed(ltop)
path(ltop)
chunkdim(ltop)
is_sparse(ltop)
```

We can now use BiocSingular to run the actual PCA. We use `deferred=TRUE` to defer centering and avoid breaking sparsity.

```{r pca}
system.time(random_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod}
system.time(random_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

Here we apply BiocSklearn to the ltop matrix. Does not seem to work with TENx files, maybe because the dataset is "assumed to be 2-dimensional array"?


```{r biocsklearn, eval=FALSE}
library(BiocSklearn)
ltop_path <- path(ltop)

system.time(sk_pca <- skIncrPCA_h5(ltop_path, n_components=50,
                                   dsname = seed(ltop)@group, chunk.size = 5000))
```

Here we apply the python methods.

```{python py_pca}
import scanpy as sc
import anndata as ad
from sklearn.decomposition import PCA, IncrementalPCA
import time

sc.settings.verbosity  = 3

filename = ltop_path
adata = sc.read_10x_h5(filename)

## scanpy PCA
sc.tl.pca(adata)

## scikitlearn incremental PCA
tic = time.perf_counter()
ipca = IncrementalPCA(n_components=50)
X_ipca = ipca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic

tic = time.perf_counter()
pca = PCA(n_components=50)
X_pca = pca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic
```

# Dense HDF5 file (TENxBrainData)

First, we read the data in dense format from the TENxBrainData package.

```{r readin_dense}
sce <- TENxBrainData()
sce
```

For consistency, we use log transformation.

```{r log_dense}
assay(sce, "logcounts") <- log1p(counts(sce))
```

We select the top variable genes. For consistency, we use a random subset of cell to compute the variances.

```{r var_dense}
set.seed(134)
idx <- sample(seq_len(ncol(sce)), 1e4)
idx <- sort(idx)
system.time(vars <- rowVars(logcounts(sce[,idx])))
names(vars) <- rownames(sce)
vars <- sort(vars, decreasing = TRUE)

geneidx <- which(rownames(sce) %in% names(vars)[1:1000])
top <- sce[geneidx,]
is_sparse(logcounts(top))
```

Log and subset are delayed operations, we need to realize the matrix to be able to run PCA.

```{r realize_dense}
ltop_dense <- realize(logcounts(top), BACKEND = "HDF5Array")
ltop_dense <- realize(t(ltop_dense), BACKEND = "HDF5Array")
seed(ltop_dense)
path(ltop_dense)
chunkdim(ltop_dense)
is_sparse(ltop_dense)
```

We can now use BiocSingular to run the actual PCA. We don't need `deferred=TRUE` here, since we are not using sparsity, but I'll keep it for consistency.

```{r pca_dense}
system.time(random_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod_dense}
system.time(random_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

Here we apply BiocSklearn to the ltop matrix.

```{r biocsklearn_dense}
ltop_path <- path(ltop_dense)
system.time(sk_pca <- skIncrPCA_h5(ltop_path, n_components=50,
                                   dsname = seed(ltop)@group, chunk.size = 5000))
```

Here we apply the python methods.

```{r write_h5ad}
library(zellkonverter)
sce <- SingleCellExperiment(assays = list(counts = ltop_dense))
writeH5AD(sce, file = "top_dense.h5ad")
```

```{python py_pca_dense}
import scanpy as sc
import anndata as ad
from sklearn.decomposition import PCA, IncrementalPCA
import time

sc.settings.verbosity  = 3

filename = "top_dense.h5ad"
adata = sc.read_h5ad(filename)

## scanpy PCA
sc.tl.pca(adata)

## scikitlearn incremental PCA
tic = time.perf_counter()
ipca = IncrementalPCA(n_components=50)
X_ipca = ipca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic

tic = time.perf_counter()
pca = PCA(n_components=50)
X_pca = pca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic
```

# Sparse TileDb

First, we transform the data in a TileDBArray.

```{r transform}
library(TileDBArray)
ltop_tile <- as(ltop, "TileDBArray")
ltop_tile

is_sparse(ltop_tile)
```

We can now use BiocSingular to run the actual PCA. We don't need `deferred=TRUE` here, since we are not using sparsity, but I'll keep it for consistency.

```{r pca_dense}
system.time(random_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod_dense}
system.time(random_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(ltop_tile, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

