---
title: "Compare PCA on TileDB sparse file"
author: "Davide Risso"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="markup")
```

```{r packages, message=FALSE, warning=FALSE}
library(TileDBArray)
library(BiocSingular)
library(BiocParallel)

ncores <- 6
```

# Sparse TileDb

First, we read the data in a TileDBArray.

```{r transform}
tenx <- TileDBArray("1M_neurons_data_sparse.tdb")
tenx <- t(tenx)
class(tenx)
is_sparse(tenx)
```

We can now use BiocSingular to run the actual PCA. We don't need `deferred=TRUE` here, since we are not using sparsity, but I'll keep it for consistency.

```{r pca_dense}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

head(random_pca$x[,1:2])
head(exact_pca$x[,1:2])

```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod_dense}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
#                                                                                  fold = 1),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

head(random_pca$x[,1:2])
head(exact_pca$x[,1:2])

```

