---
title: "Compare PCA on HDF5 sparse file (10X format)"
author: "Davide Risso"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="markup")
```

```{r packages, message=FALSE, warning=FALSE}
library(HDF5Array)
library(BiocSingular)
library(BiocParallel)

ncores <- 6
```

# Sparse HDF5 file (TENxMatrix)

First, we read the data in sparse format from the 10X Genomics file.

```{r readin}
tenx <- TENxMatrix("1M_neurons_data_sparse.h5")
class(tenx)
is_sparse(tenx)
tenx <- t(tenx)
```

We use BiocSingular to run the PCA. We use `deferred=TRUE` to defer centering and avoid breaking sparsity.

```{r pca}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

Here we apply BiocSklearn to the tenx matrix. Does not seem to work with TENx files, maybe because the dataset is "assumed to be 2-dimensional array"?


```{r biocsklearn, eval=FALSE}
library(BiocSklearn)
ltop_path <- path(tenx)

system.time(sk_pca <- skIncrPCA_h5(ltop_path, n_components=50,
                                   dsname = seed(tenx)@group, chunk.size = 5000))
```

Here we apply the python methods.

```{python py_pca}
import scanpy as sc
import anndata as ad
from sklearn.decomposition import PCA, IncrementalPCA
import time

sc.settings.verbosity  = 3

filename = ltop_path
adata = sc.read_10x_h5(filename)

## scanpy PCA
sc.tl.pca(adata)

## scikitlearn incremental PCA
tic = time.perf_counter()
ipca = IncrementalPCA(n_components=50)
X_ipca = ipca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic

tic = time.perf_counter()
pca = PCA(n_components=50)
X_pca = pca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic
```
