---
title: "Compare PCA on HDF5 dense file (TENxBrainData package)"
author: "Davide Risso"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="markup")
```

```{r packages, message=FALSE, warning=FALSE}
library(HDF5Array)
library(BiocSklearn)
library(BiocSingular)
library(BiocParallel)
library(scuttle)

ncores <- 6
```


# Dense HDF5 file (TENxBrainData)

First, we read the data in dense format from the TENxBrainData package.

```{r readin_dense}
tenx <- TENxMatrix("1M_neurons_data_dense_t.h5")
class(tenx)
is_sparse(tenx)
```

We use BiocSingular to run the actual PCA. 

```{r pca_dense}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::ExactParam(),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

head(random_pca$x[,1:2])

```

We don't need `deferred=TRUE` here, since we are not using sparsity, but I'll keep it for consistency.

```{r pca_def}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = FALSE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

head(random_pca$x[,1:2])
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod_dense}
system.time(random_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(exact_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
#                                                                                  fold = 1),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

# system.time(irlba_pca <- BiocSingular::runPCA(ltop_dense, rank = 50,
#                                               center = TRUE, scale = FALSE,
#                                               BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
#                                                                                  fold = 1),
#                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

head(random_pca$x[,1:2])

```

Here we apply BiocSklearn to the ltop matrix.

```{r biocsklearn_dense}
system.time(sk_pca <- skIncrPCA_h5("1M_neurons_data_dense_t.h5", n_components=50,
                                   dsname = "mm10", chunk.size = 5000))
```

Here we apply the python methods.

```{python py_pca_dense}
import scanpy as sc
from sklearn.decomposition import PCA, IncrementalPCA
import time

sc.settings.verbosity  = 3

filename = "1M_neurons_data_dense_t.h5"
name = "mm10"
adata = sc.read_hdf(filename, name)

## scanpy PCA
sc.tl.pca(adata)

## scikitlearn incremental PCA
tic = time.perf_counter()
ipca = IncrementalPCA(n_components=50)
X_ipca = ipca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic

tic = time.perf_counter()
pca = PCA(n_components=50)
X_pca = pca.fit_transform(adata.X)
toc = time.perf_counter()
toc - tic
```

