---
title: "Compare PCA on HDF5 sparse file (10X format)"
author: "Davide Risso"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results="markup")
```

```{r packages, message=FALSE, warning=FALSE}
library(HDF5Array)
library(BiocSingular)
library(BiocParallel)

ncores <- 6
```

# Sparse HDF5 file (TENxMatrix)

First, we read the data in sparse format from the 10X Genomics file.

```{r readin}
tenx <- TENxMatrix("1M_neurons_data_sparse.h5")
class(tenx)
is_sparse(tenx)
tenx <- t(tenx)
```

We use BiocSingular to run the PCA. We use `deferred=TRUE` to defer centering and avoid breaking sparsity.

```{r pca}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```

From the BiocSingular documentation, "f the dimensions of the input matrix are very different, it may be faster to compute the cross-product and perform the SVD on the resulting square matrix". Here we try that route.

```{r crossprod}
system.time(random_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                               center = TRUE, scale = FALSE,
                                               BSPARAM = BiocSingular::RandomParam(deferred = TRUE,
                                                                                   fold = 1),
                                               BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(exact_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::ExactParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))

system.time(irlba_pca <- BiocSingular::runPCA(tenx, rank = 50,
                                              center = TRUE, scale = FALSE,
                                              BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE,
                                                                                 fold = 1),
                                              BPPARAM = BiocParallel::MulticoreParam(ncores)))
```
